A Toy Analytic Transit Survey: Varying the Light Ratio of Binaries

### 2017-08-30 memos/toy_analytic_survey_vary_lightratio_170803.tex
initial commit

In this memo $\gamma_R$ varies across the population of star systems.
It does so because the underlying mass ratio varies, and $\gamma_R = q^\alpha$. 
Since we are interested in solar type binaries, we take the distribution of the 
mass ratio $q=M_2/M_1$ by coarsely approximating [Rhagavan et al 2010, Fig 16]:
\begin{equation}
{\rm prob}(q) =
\Bigg\{\begin{array}{lr}
c_q & 0.1 < q \leq 1  \\
0 & {\rm otherwise},
\end{array}
\label{eq:mass_ratio}
\end{equation}
for $c_q = 1/9$ to normalize the distribution to 1.
The mass-luminosity relation is that given by [Salaris \& Cassisi (2005), 
Sec 5.7],
\begin{equation}
L \propto
\Bigg\{\begin{array}{lr}
M^{4.5} & 0.5 < M_\star/M_\odot \leq 2.0  \\
M^{2.6} & 0.2 < M_\star/M_\odot \leq 0.5.
\end{array}
\label{eq:mass_luminosity}
\end{equation}
For $M_1 = M_\odot$, this gives $\alpha = 4.5$ for $0.5 < q \leq 1$, and 
otherwise $\alpha = 2.6$. 
For analytic convenience this distribution may also be approximated as 
a constant, \textit{e.g.,} $\alpha = 3$, in the subsequent development.

@@

\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/gammaR_distribn_vol_limited.pdf}
	\end{center}
	\caption{The distribution of the luminosity ratio for a volume limited 
	sample of binary stars. ${\rm BF} = 0.45$ [Duchene and Kraus 2013]; total 
	number density from Bovy 2017; $M(L)$ relation used is 
	Eq.~\ref{eq:mass_luminosity}; mass ratios are drawn from 
	Eq.~\ref{eq:mass_ratio}.}
	\label{fig:gammaR_distribn_vol_limited}
\end{figure}
\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/q_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the mass ratio for a magnitude limited 
		sample of binary stars. ${\rm BF} = 0.45$ [Duchene and Kraus 2013]; 
		total 
		number density from Bovy 2017; $M(L)$ relation used is 
		Eq.~\ref{eq:mass_luminosity}; mass ratios are drawn from 
		Eq.~\ref{eq:mass_ratio} -- a \textit{uniform} distribution in a 
		volume-limited sample!}
	\label{fig:q_distribn_mag_limited}
\end{figure}
\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/gammaR_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the luminosity ratio for a magnitude limited 
		sample of binary stars. ${\rm BF} = 0.45$ [Duchene and Kraus 2013]; 
		total 
		number density from Bovy 2017; $M(L)$ relation used is 
		Eq.~\ref{eq:mass_luminosity}; mass ratios are drawn from 
		Eq.~\ref{eq:mass_ratio}.}
	\label{fig:gammaR_distribn_mag_limited}
\end{figure}

@@

We return to the original question: how many stars?
\begin{equation}
\langle N_{\rm stars} \rangle = N_s + 2 \langle N_d \rangle,
\end{equation}
as before.
Since we are not giving analytic expressions for either of the $\langle \ldots 
\rangle$ terms, we just note that in the case of a population with a single 
$\gamma_R$ value, $N_d/N_s = {\rm BF} \times (1 + \gamma_R)^{3/2}$. A lower
bound for this ratio will always be the binary fraction.
For a single run of a Monte Carlo code, where ${\rm mean}(\gamma_R) = 0.30$, 
${\rm median}(\gamma_R) = 0.16$, and the distribution was that given in 
Fig.~\ref{fig:gammaR_distribn_mag_limited}, the observed ratio was $N_d/N_s = 
0.62$.
This corresponds to a single-valued population with $\gamma_R \sim 0.23$, a
number between the mean and median of the true distribution.
It means $\sim 1.2$ stars in binary star systems for every star in a single 
star system from this sample.

We also note in passing that $N_d/N_s = 0.62$ is a smaller ratio than for the 
case of a fixed $\gamma_R$ population with $\gamma_R = 1$, which gave $N_d/N_s 
= 1.27$. The difference is that in the latter population, the volume of 
searchable stars is greater. Based on the fixed-$\gamma_R$ population scaling 
law $N_d / N_s \propto (1+\gamma_R)^{3/2}$ (derived in the first memo), we 
would expect the ratio of single stars to be $\approx (2/1.1)^{3/2} = 2.45$ 
times smaller, which is roughly (though not exactly) what we observe.

@@

For the nominal G2V dwarf case of ${\rm BF}=0.45$, and the same distributions 
used to generate 
Figs.~\ref{fig:gammaR_distribn_vol_limited}~--~\ref{fig:gammaR_distribn_mag_limited},
we found $N_d/N_s=0.62$.
We are assuming $\Gamma_{t,d} / \Gamma_{t,s} = 1$.
If we maintained the (incorrect) assumption that $f_{d,{\rm geom}} / f_{s,{\rm 
geom}} = 1$, this would give a misclassification rate of 45\%.

@@

Computing it all out in a numerical simulation, the only thing that changes 
analytically is that
\begin{equation}
N_d f_d \rightarrow \sum_{i=1}^{N_d} f_{d,i},
\end{equation}
so when considering the ratio $f_d/f_s$, we are in fact interested in
\begin{equation}
\frac{\langle f_d \rangle}{f_s} = \frac{\frac{1}{N_d} \sum_{i=1}^{N_d} 
f_{d,i}}{f_s},
\end{equation}
where $f_{d,i}$ is the vector of geometric transit probabilities across all 
systems.
For the same numerical realization with $N_d/N_s = 0.62$, and using the mass 
radius relation given by Eq.~\ref{eq:mass_radius}, I get $\langle f_d 
\rangle / f_s = 0.84$ (smaller secondaries lead to a smaller average transit 
probability in double star systems), for a radius misclassification rate of 
$0.51$.

@@

For our nominal case, this means a relative error of $62\%$.
As discussed in Sec.~\ref{sec:how_many_stars}, this is less than our earlier 
(twin binary model) case because of the smaller volume of searchable binary 
stars.

>>>>

In this memo $\gamma_R$ varies across the population of star systems.
It does so because the underlying mass ratio varies, and $\gamma_R = q^\alpha$. 
Since we are interested in solar type binaries, we take the distribution of the 
mass ratio $q=M_2/M_1$ by coarsely approximating [Rhagavan et al 2010, Fig 16]:
\begin{equation}
{\rm prob}(q) =
\Bigg\{\begin{array}{lr}
c_q & 0.1 < q \leq 1  \\
0 & {\rm otherwise},
\end{array}
\label{eq:mass_ratio}
\end{equation}
for $c_q = 1/9$ to normalize the distribution to 1.
The mass-luminosity relation can, for analytic convenience, be approximated as 
$L = M^\alpha$, with the lore-value of $\alpha$ being $3.5$.
While we use this in subsequent analytic development, for numerics we fit a 
lines to mass-luminosity data collected by Torres et al. [2009] for 
dwarfs above M, and Benedict et al. [2016] for dwarfs below M.
To convert Benedict et al. [2016]'s reported $M_V$ values to absolute 
luminosities, we interpolated over E. Mamajek's 
table\footnote{
\texttt{pas.rochester.edu/\textasciitilde 
emamajek/EEM\_dwarf\_UBVIJHK\_colors\_Teff.txt},
downloaded 2017.08.02}.
In log-log space, we let the intersection point of the lines float, and make 
various cuts on the data as indicated in Fig~\ref{fig:mass_luminosity}. The 
fit parameters are available in a footnote\footnote{
	m lo: 1.8818719873988132~--~
	c lo: -0.9799647314108376~--~
	m hi: 5.1540712426599882~--~
	c hi: 0.0127626185389781~--~
	M at merge: 0.4972991257826812~--~
	L at merge: 0.0281260412126928.
	}.
\begin{figure}
	\begin{center}
		\includegraphics[scale=0.9]{figures/mass_luminosity.pdf}
	\end{center}
	\caption{Empirical fit to main sequence dwarf mass luminosity data from 
	Torres et al. [2009] and Benedict et al. [2016]. The ``low fit'' is a least 
	squares fit to data from $0.1-0.5M_\odot$, and the ``high fit'' is to data 
	above that, and below the Kraft break ($1.2M_\odot$).
	The $L(M)$ relation taken for subsequent numerics is the maximum 
	of the two fits.
	}
	\label{fig:mass_luminosity}
\end{figure}

@@

In passing, the bias in ${\rm prob}(\gamma_R)$ towards low luminosity ratios 
can be seen analytically. If we assume $\gamma_R = q^\alpha$, then
\begin{equation}
{\rm prob}(\gamma_R) = \frac{1}{9 \alpha} \ \gamma_R^{\frac{1-\alpha}{\alpha}}
\quad {\rm for\ } (0.1)^\alpha < \gamma_R < 1,
\end{equation}
and otherwise zero. For instance if $\alpha = 3$, ${\rm prob}(\gamma_R) 
\propto \gamma_R^{-2/3}$ and the domain extended from $1$ to 
$10^{-3}$, where the probability distribution peaks.

\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/gammaR_distribn_vol_limited.pdf}
	\end{center}
	\caption{The distribution of the luminosity ratio for a volume limited 
	sample of binary stars. ${\rm BF} = 0.45$ [Duchene and Kraus 2013]; total 
	number density from Bovy 2017; $M(L)$ relation from 
	Fig.~\ref{fig:mass_luminosity}; mass ratios are drawn from 
	Eq.~\ref{eq:mass_ratio}.}
	\label{fig:gammaR_distribn_vol_limited}
\end{figure}
\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/q_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the mass ratio for a magnitude limited 
		sample of binary stars. ${\rm BF} = 0.45$ [Duchene and Kraus 2013]; 
		total 
		number density from Bovy 2017; $M(L)$ relation from 
		Fig.~\ref{fig:mass_luminosity}; mass ratios are drawn from 
		Eq.~\ref{eq:mass_ratio} -- a \textit{uniform} distribution in a 
		volume-limited sample!}
	\label{fig:q_distribn_mag_limited}
\end{figure}
\begin{figure}[!t]
	\begin{center}
		\includegraphics[scale=.8]{figures/gammaR_distribn_mag_limited.pdf}
	\end{center}
	\caption{The distribution of the luminosity ratio for a magnitude limited 
		sample of binary stars. ${\rm BF} = 0.45$ [Duchene and Kraus 2013]; 
		total 
		number density from Bovy 2017; $M(L)$ relation from 
		Fig.~\ref{fig:mass_luminosity}; mass ratios are drawn from 
		Eq.~\ref{eq:mass_ratio}.}
	\label{fig:gammaR_distribn_mag_limited}
\end{figure}

@@

We return to the original question: how many stars?
\begin{equation}
\langle N_{\rm stars} \rangle = N_s + 2 \langle N_d \rangle,
\end{equation}
as before.
Since we are not giving analytic expressions for either of the $\langle \ldots 
\rangle$ terms, we just note that in the case of a population with a single 
$\gamma_R$ value, $N_d/N_s = {\rm BF} \times (1 + \gamma_R)^{3/2}$. A lower
bound for this ratio will always be the binary fraction.
For a single run of a Monte Carlo code, where ${\rm mean}(\gamma_R) = 0.30$, 
${\rm median}(\gamma_R) = 0.16$, and the distribution was that given in 
Fig.~\ref{fig:gammaR_distribn_mag_limited}, the observed ratio was $N_d/N_s = 
0.59$.
This corresponds to a single-valued population with $\gamma_R \sim 0.23$, a
number between the mean and median of the true distribution.
It means $\sim 1.2$ stars in binary star systems for every star in a single 
star system from this sample.

We also note in passing that $N_d/N_s = 0.59$ is a smaller ratio than for the 
case of a fixed $\gamma_R$ population with $\gamma_R = 1$, which gave $N_d/N_s 
= 1.27$. The difference is that in the latter population, the volume of 
searchable stars is greater. Based on the fixed-$\gamma_R$ population scaling 
law $N_d / N_s \propto (1+\gamma_R)^{3/2}$ (derived in the first memo), we 
would expect the ratio of single stars to be $\approx (2/1.1)^{3/2} = 2.45$ 
times smaller, which is roughly (though not exactly) what we observe.

@@

For the nominal G2V dwarf case of ${\rm BF}=0.45$, and the same distributions 
used to generate 
Figs.~\ref{fig:gammaR_distribn_vol_limited}~--~\ref{fig:gammaR_distribn_mag_limited},
we found $N_d/N_s=0.59$.
We are assuming $\Gamma_{t,d} / \Gamma_{t,s} = 1$.
If we maintained the (incorrect) assumption that $f_{d,{\rm geom}} / f_{s,{\rm 
geom}} = 1$, this would give a misclassification rate of 45\%.

@@

Computing it all out in a numerical simulation, the only thing that changes 
analytically is that
\begin{equation}
N_d f_d \rightarrow \sum_{i=1}^{N_d} f_{d,i},
\end{equation}
so when considering the ratio $f_d/f_s$, we are in fact interested in
\begin{equation}
\frac{\langle f_d \rangle}{f_s} = \frac{\frac{1}{N_d} \sum_{i=1}^{N_d} 
f_{d,i}}{f_s},
\end{equation}
where $f_{d,i}$ is the vector of geometric transit probabilities across all 
systems.
For the same numerical realization with $N_d/N_s = 0.59$, and using the mass 
radius relation given by Eq.~\ref{eq:mass_radius}, I get $\langle f_d 
\rangle / f_s = 0.84$ (smaller secondaries lead to a smaller average transit 
probability in double star systems), for a radius misclassification rate of 
$0.51$.

@@

For our nominal case, this means a relative error of $59\%$.
As discussed in Sec.~\ref{sec:how_many_stars}, this is less than our earlier 
(twin binary model) case because of the smaller volume of searchable binary 
stars.
